Annual Analysis of Delays in TTC Streetcars for the year 2023

### 2024-09-27 paper/paper.qmd
Improved wordings in report and readme

---
title: "Annual Analysis of Delays in TTC Streetcars for the year 2023"
subtitle: "Is there a way to make commuter life easier?"
author:
  - Aakash Vaithyanathan
thanks: "Code and data are available at: https://github.com/aakash2002/study_of_ttc_streetcar_delays"
date: today
date-format: long
abstract: "This paper aims to study the causes of delays in Toronto Transit Commissions (TTC) streetcar service, the variation in delays by the hours in working shifts and whether there are any seasonal relationships to the observed delays for the year 2023. By using the data provided by OpenData Toronto, we are able to understand the observed relationships between the several fields of interests like minimum delay, hour, season, streetcar line and incident type. Our study finds that there is higher reported delayed times during night shift and higher delays reported during peak hours that fall during the working 9AM - 5PM time period. Interestingly we did not observe any variation in delays during the various seasons and understand that to make stronger inferences between the temporal relationship, we need to examine longer historic data."
format: 
  pdf: 
    fig-pos: "H"
    fig-caption: true
number-sections: true
bibliography: references.bib
---

@@

These delays make it especially difficult for individuals that commute on a daily basis. Furthermore, one can image the inconveniences caused due to extended delays by streetcars during weather conditions like the winter months. In this paper, we aim to dive deeper into the cause of delays in TTC, how these vary by the times of day and the various seasons and what incidents are the most occurring for the delays experienced. The starter code for this paper was provided by Professor Rohan (@citeRohan). We make use R (@citeR) and its supported packages like tidyverse  (@tidyverse) and dplyr (@dplyr). For our dataset, we use the provided data for the year 2023 by OpenData Toronto (@opendatatoronto). The reported graphs in our study is generated using ggplot (@ggplot2) and the tables are generated using the gt (@gt) library. Code cleanup is performed using the Janitor (@janitor) package.

The remainder of this paper is structured as follows. We first give a brief overview of the dataset under [dataset description](#datasetdesc) section. Then, we present some graphs on the results extracted from our collected data found under [results](#results). We then proceed to a discussion of our results found under [discussion](#discussion) section and highlight the limitations in our conducted study found under [limitations](#limitations) section and end our study with some proposed [next steps](#nextsteps).

@@

## Overview of dataset
The following data was made available from OpenData Toronto (@opendatatoronto) for the year 2023. The data was extracted from the server and saved locally using in a raw data folder. Some of the features part of this dataset are date, time, incident, min delay, bound and location.

## Assumption made in dataset {#assumption}
We observe that our dataset has several rows where the minimum delay time is over 900 minutes or are equal to 0. These values can heavily skew the results that we may observe. Therefore, one key assumption we make in our study is that we limit the maximum delay time to upto 2 hours or 120 minutes. As a result of this filtering, we only lose about 130 data and still retain about 10723 data entries to use in our study.

@@

The raw data is undergone a series of data cleaning steps to ensure adequate usability for the study. We first use the Janitor (@janitor) package to clean the data columns removing any spaces between them and using a consistent naming stype for them. Next, we proceed to split the date field into year and month fields and split the time field extracting the hour of day from it. Following this, we create a new column to denote the season based on the month of the year from the data. This helps during the analysis stage when we need to make inferences about the seasonal variation in average delays in streetcars.

We then proceed to drop the rows with missing values or NA. Out of the total 14413 rows of data, 2399 of them contain NA values in total which doesn't heavily impact our data. Therefore, no transformations are necessary to populate missing entries in our dataset. Lastly, as highlighed in the [assumption](#assumption) section, we filter our dataset to only include delays of upto 120 minutes.

As a result of the above described data cleaning process, we select our desired features from the dataset. @tbl-dataset-preview shows the first five rows of our data and the selected features.

@@

The dataset was last updated September 24, 2024 ensuring consistent updates to the collected data. OpenData Toronto also provides functionality and sufficient instructions for developers to easily query their data using scripts in Python, R and Node.js instead of having to manually download them.

The website doesn't provided information on how the data was sourced but we can assume that the data is recorded by taking the difference in the reported arrival time via the TTC app and the actual time the streetcar arrives at various stops. The collected data represents common types like characters, numeric and datetime. Therefore, for the purposes of this study, the procedure and type of the data reported is appropriate for our use and we do not require any extra information to make any assumptions about the data types.

@tbl-selected-features describes the data types for the selected features in our study from the cleaned dataset after the necessary transformations.

# Results {#results}
## Average Delay By Streetcar Line
```{r}
#| include: true
#| warning: false
#| message: false
#| echo: false
#| label: fig-delay-by-line
#| fig-cap: Average annual delay in minutes for each streetcar line

@@

@fig-delay-by-line highlights the annual average delay in minutes for every streetcar line for the year 2023. The 300-series lines also called the Blue Network Routes (@ttcRoutesSchedule) are the nighttime streetcar lines that start at 1:30 AM. These buses average the most delays reaching at most 30 minutes for line 312.

The 500-series lines are the daytime lines that run from 8 AM. These buses average between 6 to 23.6 minutes with the longest delay experienced by line 507.

## Average Delays During Dayshift
```{r}
#| label: fig-delay-dayshift
#| fig-cap: Average delay in minutes for dayshift
#| echo: false
#| warning: false
#| message: false

@@

## Average Delays During Nightshift
```{r}
#| label: fig-delay-nightshift
#| fig-cap: Average delay in minutes for nightshift
#| echo: false
#| warning: false
#| message: false

@@

# Plot with percentage labels
ggplot(freq_incidence, aes(x = incident, y = frequency)) +
  geom_bar(stat = "identity", fill = "dodgerblue1", alpha = 0.75) +
  geom_text(aes(label = paste0(round(percentage, 1), "%")), 
            vjust = -0.5,
            size = 2) +
  ggtitle("Frequency and percentage of incident types in 2023") +
  xlab("Delay Incident Type") +
  ylab("Frequency") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),  # Center title
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7)  # Rotate x-axis labels for clarity
  )
```
@fig-freq-delay-incidents highlights the total annual delays incurred based on incidents reported by TTC for the reason. We can see for the year of 2023, 23% of delay incidents were due to operation delays amounting to a total of 2463 instances and rail switches were the least occurring with reported instances being only about 0.2% of all delays.

## Delay by Seasons
```{r}
#| label: fig-delay-by-season
#| fig-cap: Average delay in minutes by seasons in a year
#| echo: false
#| warning: false
#| message: false

@@

## Delays by bus lines
The results from @fig-delay-by-line shows that blue network routes (@ttcRoutesSchedule) buses incur the highest average delay. This result makes sense as compared to the day shift buses, there are far fewer night shift buses resulting in fewer drivers for various locations the buses may take. Additionally, we also note that the 900-series streetcar line that go to the Airport incur the longest average delay of almost 54 minutes.

## Average delay for morning shift
The results from @fig-delay-dayshift showcases the variability in average delays by the hour during the morning shift working hours observed between 8 AM to 6 PM. We observe that the results are in-line with the expected behavior of longer average delays of around 18 minutes during peak hours like 8 AM or 5 PM which is usually when a large working population leave and return from work. This can result in higher traffic hours on average causing an expected delay in streetcar arrivals. Additionally, we notice that throughout the day there are several points of interest for peak delays as seen at 11 AM or 1 PM incurring an average delay of 19 minutes. This observation could indicate the need for travel to various locations like grocery stores, perhaps to a friend's house or classes for university students. During these hours, there is heavy traffic and these suggest a potential reason for delays.

## Average delay for night shift
The results from @fig-delay-nightshift shows a decreasing trend in the average delays by the hour during the night shift working hours observed between 2 AM to 7 AM. These findings are what we might expect as there is fewer traffic delays during the night resulting in a more consistent decreasing delay pattern. A possible reason for longer delay duration despite fewer traffic could be the fact there there are far fewer blue netwrok routes (@ttcRoutesSchedule) buses that run resulting in fewer drivers that require to cover longer distances. The graph shows that as we get later into the night shift, we notice the delays dropping as low as roughly 14 minutes.

## Annual frequency of reported delay incidents
The results from @fig-freq-delay-incidents showcase the total count of instances of delays by various incident reports in the year 2023. The above bar graph shows that the most occurring incident was due to operational delays accounting for 2467 total reports. The least occurring incident was Rail switches accounting for only 21 total instance reports in the year.

## Annual delay in minutes by seasons in a year
The results from @fig-delay-by-season show that on average there wasn't as much variability in delays between the various seasons in a year with summer month accounting for the longest reported average delays. Interestingly, the winter month accounted to the least average delays. This result is in-line with the existing literature regarding the weather reports in Canada during the winter season as for the first time in history, Toronto reported an average temperature above -5 degrees Celsius throughout the month of December and also reported a warmer winter season overall (@canadaWinterArticle).

## Conclusion
By using the findings from @fig-delay-by-line, and @fig-delay-nightshift, we can conclude that there on average the night shift streetcars showcased the longest average delays with periods between 2 AM to 4 AM having the most delays. These findings support our understanding that there are far fewer night shift buses that operate on average compared to the number of streetcars that operate during the day's regular working hours. @fig-delay-dayshift reports that the day shift streetcars have their reported delays being high during peak hours like 8 AM, 11 AM, 1 PM, 5 PM or 8 PM. These results are in-line with the idea that hours between 9-5 are the general working hours in a day and likely can result in high traffic volumes for individuals and commuters.

@@

The results from our study highlight important considerations on the types of delay incidents that occur in streetcars, what time of day between various working shifts are there highest reported delays and whether season has any impact on the cause of delays. By examining these results, we hope to be able to allocate improved resources to address the various reported incidents for delays. To address the issue of delays during peak hours, a potential solution might be to employ additional streetcars that run during these hours and by investing more resources in the night shift streetcars the workload and reduce long delays during late hours like 2-4 AM as these hours are usually dangerous to be out especially alone.

@@

In terms of the next steps, there are 2 ways we could extend our study to improve the findings from it.

The first is by addressing the above mentioned limitation of combining data from various years to provide a more richer quality of identified results. By combining the data from past years, we are better able to learn of any temporal trends in the data which can lead to interesting results.

The second way to improve our study is by making use of available streetcar delay sentiment data to understand the recurring pain points and emotions that are expressed by commuters. This data when combined with existing streetcar delay data can provide richer contextual information about the types of incident delays and the sentiment expressed. Another interesting result we can hope to learn is how these sentiments vary between various seasons as during harsher seasons like summer and winter, one can imagine that it wouldn't be most pleasant to experience a long delay in streetcar arrivals.

These findings can help to better connect with the population and provide valuable information on how to better modify or establish new policies that can help to improve the overall commuting experience for the residents of Toronto, the city they call home.

>>>>

---
title: "Annual Analysis of Delays in TTC Streetcars for the year 2023"
subtitle: "Is there a way to make commuter life easier?"
author:
  - Aakash Vaithyanathan
thanks: "Code and data are available at: https://github.com/aakash2002/study_of_ttc_streetcar_delays"
date: today
date-format: long
abstract: "This paper aims to study the causes of delays in Toronto Transit Commissions (TTC) streetcar service by examining the relationship between various factors like hour of day, streetcar lines and seasons. Our study found that the night shift had the higest delays by average of 25 minutes with high variability in delays observed during the hours of 9 AM to 5 PM. Interestingly, we did not observe any seasonal variations in the overall delay in streetcar and one possible reason for this could be due to our limited dataset. We beleive the results from this paper can help in better designing policies and transit schedules to improve the overall commuting experience."
format: 
  pdf: 
    fig-pos: "H"
    fig-caption: true
number-sections: true
bibliography: references.bib
---

@@

The delays via streetcars make it especially difficult for commuters on a daily basis. One can image the inconveniences by these delays during the harsh winter weather conditions and the struggles commuters face. In 2019, streetcars averaged delays of about 13,790 minutes each month and by 2023, this number increased by 33% (@streetcarDelay2023). As a result, commuters are in need for improved policies and better transit schedules to address the difficulties they face everyday. Our study found that the night shift had the higest delays by average of 25 minutes with high variability in delays observed during the hours of 9 AM – 5 PM. Interestingly, we did not observe any seasonal variations in the overall delay in streetcar and one possible reason for this could be due to our limited dataset.

In this paper, we aim to dive deeper into the cause of delays in TTC, how these vary by the times of day and the various seasons and what incidents are the most occurring for the delays experienced. The starter code for this paper was provided by Professor Rohan (@citeRohan). We make use R (@citeR) and its supported packages like tidyverse  (@tidyverse) and dplyr (@dplyr). For our dataset, we use the TTC streetcar delay data for the year 2023 by OpenData Toronto (@opendatatoronto). The reported graphs in our study are generated using ggplot (@ggplot2) and the tables are generated using the gt (@gt) package. The process to perform data cleaning was done using the Janitor (@janitor) package.

The remainder of this paper is structured as follows. We first give a brief overview of the dataset under [dataset description](#datasetdesc) which talks about the various features, the assumptions we make and the measurements done to gather the data. We present some graphs on the results extracted from our collected data found under [results](#results). A discussion of our results can be found under [discussion](#discussion) section with the relevant limitations in our study found under [limitations](#limitations) section. We end our study with some proposed [next steps](#nextsteps) for future version of this paper.

@@

## Overview of dataset
The following data was made available from OpenData Toronto (@opendatatoronto) for the year 2023. The data was extracted from the server and saved locally using in a raw data folder using the instructions provided by the documentation. We note that this was the only available dataset with frequent updates and data dating back to 2014. Some of the features part of this dataset are date, time, incident, min delay, bound and location.

## Assumption made in dataset {#assumption}
We observe that our dataset has several rows where the minimum delay time is over 900 minutes or are equal to 0. These values can heavily skew the results that we may observe. Therefore, one key assumption we make in our study is that we are only interested in at most a delay by 2 hours in streetcars. Therefore, during the data cleaning step, we will limit the rows of data to 120 minutes. As a result of this filtering, we only lose about 130 data and still retain about 10723 data entries to use in our study.

@@

The raw data is undergone a series of data cleaning steps to ensure adequate usability for the study. We first use the Janitor (@janitor) package to clean the data columns removing any spaces between them and using a consistent variable name convention. Next, we proceed to split the date field into year and month fields and do the same with the time field extracting the hour of day from it. Following this, we create a new column to denote the season based on the month of the year from the data. This is used for studying any seasonal variation in delays that may exist in our data.

After selecting the appropriate columns in our dataset, we proceed to drop the rows with missing values in them. Out of the total 14413 rows of data, 2399 of them contain missing values. As a consequence of dropping these rows, we do not lose any valuable information in our dataset. Therefore, no transformations are necessary to populate any missing entries. Lastly, as highlighted in the [assumption](#assumption) section, we filter our data to only include delays up to 120 minutes.

@tbl-dataset-preview shows the first five rows of our data and the selected features.

@@

The dataset was last updated September 24, 2024 ensuring consistent updates to the collected data. OpenData Toronto also provides functionality and sufficient instructions for developers to easily query their data using scripts in Python (@python), R (@citeR) and Node.js (@nodeJS) instead of having to manually download them.

The website doesn't provided information on how the data was sourced but we can assume that the data is recorded by taking the difference in the reported arrival time via the TTC app and the actual time the streetcar arrives at various stops. The collected data represents common types like characters, numeric and datetime. Therefore, for the purposes of this study, the procedures used to collect the data and their types are appropriate for our use and we do not require any extra information to make any assumptions them.

@tbl-selected-features describes the data types for the selected features in our study from the cleaned dataset.

# Results {#results}
## Average Delay By Streetcar Line
```{r}
#| include: true
#| warning: false
#| message: false
#| echo: false
#| label: fig-delay-by-line
#| fig-cap: Average annual delay in minutes for each streetcar line in 2023

@@

@fig-delay-by-line highlights the annual average delay in minutes for every streetcar line for the year 2023. The 300-series lines also called the Blue Network Routes (@ttcRoutesSchedule) are the night time streetcar lines that start at 12:00 AM. These average the most delays reaching at most 30 minutes for line 312.

The 500-series lines are the daytime lines that run from 8 AM. These streetcar average between 6 to 23.6 minutes with the longest delay experienced by line 507.

## Average Delays During Dayshift
```{r}
#| label: fig-delay-dayshift
#| fig-cap: Average delay in minutes for dayshift in 2023
#| echo: false
#| warning: false
#| message: false

@@

## Average Delays During Nightshift
```{r}
#| label: fig-delay-nightshift
#| fig-cap: Average delay in minutes for nightshift in 2023
#| echo: false
#| warning: false
#| message: false

@@

# Plot with percentage labels
ggplot(freq_incidence, aes(x = incident, y = frequency)) +
  geom_bar(stat = "identity", fill = "dodgerblue1", alpha = 0.75) +
  geom_text(aes(label = paste0(round(percentage, 1), "%")), 
            vjust = -0.5,
            size = 2) +
  ggtitle("Frequency and percentage of incident types in 2023") +
  xlab("Delay Incident Type") +
  ylab("Frequency") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),  # Center title
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7)  # Rotate x-axis labels for clarity
  )
```
@fig-freq-delay-incidents highlights the total annual delays incurred based on incidents reported by TTC for the cause of said delay. We can that 23% of delay incidents were due to operational reasons amounting to a total of 2463 instances and rail switches were the least occurring with reported instances being only about 0.2% of all delays.

## Delay by Seasons
```{r}
#| label: fig-delay-by-season
#| fig-cap: Average delay in minutes by seasons in 2023
#| echo: false
#| warning: false
#| message: false

@@

## How do delays vary by streetcars lines?
The results from @fig-delay-by-line shows that blue network routes (@ttcRoutesSchedule) streetcar incur the highest average delay of about 25 minutes. This result makes sense as compared to the day shift streetcar, there are far fewer night shift streetcar resulting in fewer drivers for various locations the streetcar may take.

## Are night time streetcars busier than the day?
The results from @fig-delay-dayshift showcases the variability in average delays by the hour during the morning shift working hours observed between 8 AM to 6 PM. We observe that the results are in-line with the expected behavior of longer average delays of around 18 minutes during peak hours like 8 AM or 5 PM which is usually when a large working population leave for and return from work. This can result in higher traffic hours on average causing an expected delay in streetcar arrivals. Additionally, we notice that throughout the day there are several points of interest for peak delays as seen at 11 AM or 1 PM incurring an average delay of 19 minutes. This observation could indicate the need for travel to various locations like grocery stores, perhaps to a friend's house or university students commuting for their class.

The results from @fig-delay-nightshift show the average delays by the hour during the night shift working hours observed between 12:00 AM to 7 AM. These findings are what we might expect as there is fewer traffic delays during the night resulting in a more consistent decreasing delay pattern.

Despite daytime being busier than nighttime, why do streetcars at night report the highest reported delays? A possible reason for longer delays despite fewer traffic could be the fact there there are far fewer blue network routes (@ttcRoutesSchedule) streetcar that run resulting in fewer drivers that require to cover longer distances. The graph shows that as we get later into the night shift, we notice the delays dropping as low as 11 minutes.

## What causes the delays in streetcars?
The results from @fig-freq-delay-incidents showcase the total count of instances of delays by various incident reports in the year 2023. This graph shows that the most occurring incident was due to operational delays accounting for 23% of total reports with rail switches being the least amounting to just 0.2 %.

## Is it true that winters making commuting especially challenging?
The results from @fig-delay-by-season show that on average there wasn't as much variability in delays between the various seasons in a year with summer month accounting for the longest reported average delays. We find it interesting that winter accounted to the least average delays as in 2015, 25 streetcars were knocked off-route due to harsh weather conditions (@torontosunColdWeather). One reason for why this wasn't the case in 2023 is due to the fact that Toronto reported an average temperature above -5 degrees Celsius throughout the month of December and also reported a warmer winter season overall (@canadaWinterArticle).

## Conclusion
By utilizing the results from @fig-delay-by-line, and @fig-delay-nightshift, we can conclude that on average the night shift streetcars showcased the longest average delays with periods between 2 AM to 4 AM. These findings make sense as there are far fewer night shift streetcar that operate on average compared to the number of streetcars that operate during the day. @fig-delay-dayshift reports that the day shift streetcars have their reported delays being high during peak hours like 8 AM, 11 AM, 1 PM, 5 PM or 8 PM. These results are in-line with the idea that hours between 9 AM to 5 AM are the general working hours in a day and likely can result in high traffic volumes for commuters.

@@

The results from our study highlight important considerations on the types of delay incidents that occur in streetcars, what time of day between various working shifts are there highest reported delays and whether season has any impact on the cause of delays. By examining these results, we hope to be able to allocate resources to address the various reported incidents for delays. To address the issue of delays during peak hours, a potential solution might be to employ additional streetcars that run during these hours and by investing more resources in the night shift to reduce the long delays during late hours like 2 to 4 AM as these hours are usually dangerous to be out especially alone.

@@

In terms of the next steps, there are two ways we could extend our study to improve the findings from it.

The first is by addressing the above mentioned limitation of combining data from various years to provide a more richer quality of identified results. By combining the data from past years, we are better able to learn of any temporal relationships in the data which can lead to stronger conclusions.

The second way to improve our study is by making use of any available streetcar delay sentiment data to understand the recurring pain points and emotions that are expressed by commuters. This data when combined with existing streetcar delay data can provide richer contextual information about the types of incident delays and the sentiments expressed. Another interesting result we can hope to learn is how these sentiments vary between various seasons or times of day.

These findings can help TTC to better connect with the population and provide valuable information on how to improve policies and schedules that can help to improve the overall commuting experience for the residents of Toronto, the city they call home.
